humanCdGF_predict_cfu

### 2016-10-10 code/HumanCdGF_predict_cfu.Rmd
reduce samples to check model improvement

```{r setup data}
# read in files
meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'

# read in files
meta_file   <- read.table(meta_file, sep = '\t', header = T)
shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
source(tax_function)

# remove samples not used or with other cdiff strains
meta_file <- 
  meta_file %>% 
  filter(Early_Euth == FALSE, cdiff_strain == 431, !cage_id == 'NP1') %>% 
  select(mouse_id, sample_id, day, log_cfu) %>% 
  filter(!is.na(log_cfu))

@@

### All OTUs input to Random Forest - without mice euthanized early
```{r all otus, fig.width=10, fig.height=6}
heat.plot(rsq, '#003300', legend_title = expression(paste('R'^'2')))
#plot(otu_votes[otu_votes >= 10], xlab = 'OTU', ylab = 'Freq')
```

@@

```{r top otus, fig.width=10, fig.height=6}
heat.plot(rsq_n, '#003300', legend_title = expression(paste('R'^'2')))
#plot(otu_votes_n[otu_votes_n >= 1], xlab = 'OTU', ylab = 'Freq')
```

>>>>

```{r setup data}
# read in files
meta_data   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'

# read in files
meta_data   <- read.table(meta_data, sep = '\t', header = T)
shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
source(tax_function)

# remove samples not used or with other cdiff strains
meta_file <- 
  meta_data %>% 
  filter(Early_Euth == FALSE, cdiff_strain == 431, !cage_id == 'NP1') %>% 
  select(mouse_id, sample_id, day, log_cfu) %>% 
  filter(!is.na(log_cfu))

@@

@@

```{r top otus, fig.width=6, fig.height=6}
heat.plot(rsq_n, '#003300', legend_title = expression(paste('R'^'2')))
#plot(otu_votes_n[otu_votes_n >= 1], xlab = 'OTU', ylab = 'Freq')
```

### CFU and Rel abundance plots

```{r fig.height=5, fig.width=8}
meta_data %>% 
  filter(!is.na(day), !cage_id == 'NP1', 
         Early_Euth == FALSE, cdiff_strain == 431) %>% 
  ggplot(aes(x = factor(day), y = log_cfu, color = cage_id)) + 
    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
    theme_bw() + labs(x = 'Day', y = 'C Diff CFU (log10)') +
    theme(legend.justification=c(1,0), legend.position=c(1,0), 
          legend.title=element_blank())
```

How does removing the low samples (Day 8 and CFUs around 0) affect prediction?

```{r identify reason for lowered R2, fig.height=5, fig.width=8}
log_cutoff <- 2.5

meta_data %>% 
  filter(!sample_id == 'OUT2-2533-D8', log_cfu >= log_cutoff) %>% 
  filter(!is.na(day), !cage_id == 'NP1', 
         Early_Euth == FALSE, cdiff_strain == 431) %>% 
  ggplot(aes(x = factor(day), y = log_cfu, color = cage_id)) + 
    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
    theme_bw() + labs(x = 'Day', y = 'C Diff CFU (log10)') +
    theme(legend.justification=c(1,0), legend.position=c(1,0), 
          legend.title=element_blank()) + scale_y_continuous(limits = c(0,9))

meta_reduced <- meta_data %>% 
  filter(!sample_id == 'OUT2-2533-D8', log_cfu >= log_cutoff) %>% 
  filter(Early_Euth == FALSE, cdiff_strain == 431, !cage_id == 'NP1') %>% 
  select(mouse_id, sample_id, day, log_cfu) %>% 
  filter(!is.na(log_cfu))

rsq_r <- matrix(NA, nrow = 11, ncol = 10)
otu_votes_r <- rep(0, ncol(rel_abund_shared_01))
names(otu_votes_r) <- colnames(rel_abund_shared_01)

for(m in 0:10){
  sample_list <- shared_list_days %>% 
    filter(day == m) %>% 
    select(-day)
  community.df <- rel_abund_shared_01 %>% 
    right_join(sample_list) %>% 
    select(-sample_id)
  for(n in 1:10){
    cfu_shared.df <- meta_reduced %>% 
      filter(day == n) %>% 
      select(mouse_id, log_cfu) %>% 
      right_join(community.df) %>% 
      select(log_cfu, one_of(top_n_overall)) %>% 
      filter(!is.na(log_cfu))

    set.seed(seed)
    rf_top <- randomForest(log_cfu ~ ., data = cfu_shared.df,
                           importance = TRUE, ntree = n_trees)
    importance_sorted <- sort(importance(rf_top)[,1], decreasing = T)
    top_important_OTU <- rownames(data.frame(head(importance_sorted, n_features)))
    set.seed(seed)
    rf_n_features <- randomForest(log_cfu ~ ., data=cfu_shared.df[ ,c( 'log_cfu',
                                                                       names(importance_sorted)[1:n_features])], 
                                  importance=TRUE, ntree=n_trees)
    importance_final <- sort(importance(rf_n_features)[,1], decreasing = T)
    top_OTU_final <- rownames(data.frame(head(importance_final, n_features)))
    rsq_r[m+1, n] <- round(tail(rf_n_features$rsq, 1), 2)
    otu_votes_r <- otu_votes_r + 1*(colnames(rel_abund_shared_01) %in% top_OTU_final)
  }
}
taxa_r <- get_tax(1, names(otu_votes_r[otu_votes_r >= 1]), tax_file)
taxa_names_r <- paste0(taxa$tax, ' (', gsub('tu0*', 'TU ', rownames(taxa)),')')
```

```{r fig.width=6,fig.height=6}
heat.plot(rsq_r, '#003300', legend_title = expression(paste('R'^'2')))
```

Removing these data points improves predictability for day 6 and 8 but not the earlier days. If I increase the cutoff to remove all CFU below 4, there is still no improvement and even in some cases decreased performance.

### Is there any abnormalities in the OTUs of any these days?

```{r plot features,fig.width=8, fig.height=20}
meta_data %>% 
  filter(!sample_id == 'OUT2-2533-D8', log_cfu >= log_cutoff) %>% 
  filter(!cage_id == 'NP1', Early_Euth == FALSE, cdiff_strain == 431) %>% 
  full_join(rel_abund_shared_01) %>% 
  select(day, sample_id, mouse_id, cage_id, 
         one_of(names(otu_votes_r[otu_votes_r >= 1]))) %>% 
  gather(otu, rel_abund, Otu00002:Otu00250) %>% 
  filter(!is.na(day)) %>% 
  ggplot(aes(x=factor(day),y=rel_abund, color = cage_id)) +
    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
    facet_grid(otu ~., scales='free_y') +
    theme_bw() + labs(x = 'Day', y = 'Relative Abundance (%)')
```

***

***
